---
title: "Pstat 131 Final Project"
author: "Daniel Pry, Nick Van Daelen, Rick Zheng"
date: "2/27/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ISLR)
library(glmnet)
library(tree)
library(maptree)
library(randomForest)
library(gbm)
library(ROCR)
library(ggplot2)
```

# Pstat 131 Winter 2022 Final Project:

## Predicting if NCAA Basketball players will be drafted based on their stats.

Loading Data Set:
source: https://www.kaggle.com/adityak2003/college-basketball-players-20092021
```{r}
college_stats <- read.csv('CollegeBasketballPlayers2009-2021.csv')
college_stats %>% select(ht, pid)
```

```{r}
stats <- college_stats %>% mutate(drafted = ifelse(is.na(pick), 0, 1))
```

```{r}
s <- stats %>% group_by(pid) %>% summarise(player_name = player_name, team = team, conf = conf, seasons = length(pid), Min_per  = Min_per, Ortg  = Ortg, usg  = usg, eFG  = eFG, TS_per  = TS_per, ORB_per  = ORB_per, DRB_per  = DRB_per, TO_per  = TO_per, AST_per  = AST_per, FTM  = FTM, FTA  = FTA, FT_per = FT_per, twoPM = twoPM, twoPA = twoPA, twoP_per = twoP_per, TPM = TPM, TPA = TPA, TP_per = TP_per, blk_per = blk_per, stl_per = stl_per, college_year = yr, ht = ht, adjoe  = adjoe, year = year, pid = pid, Rec.Rank = Rec.Rank, ast.tov = ast.tov, rimmade = rimmade, rim_att = rimmade.rimmiss, rim_per = rimmade..rimmade.rimmiss., midmade = midmade, mid_att = midmade.midmiss, mid_per = midmade..midmade.midmiss., dunksmade = dunksmade, dunks_att = dunksmiss.dunksmade, dunk_per = dunksmade..dunksmade.dunksmiss., drtg = drtg, adrtg = adrtg, stops = stops, bpm = bpm, min_played = mp, off_reb = oreb, def_reb = dreb, total_reb = treb, ast = ast, stl = stl, blk = blk, pts = pts, Position = X, drafted = drafted)

s_cleaned <- s %>% slice_max(year)
head(s_cleaned)
```

```{r}
#changing ht
s_cleaned %>% select(ht)
begin_height <- data.frame(do.call("rbind", strsplit(as.character(s_cleaned$ht), "-", fixed = TRUE)))

```
<<<<<<< HEAD

```{r, fig.width=10}
begin_height["X1"][begin_height["X1"] == "Jun"] <- 72
begin_height["X1"][begin_height["X1"] == "Jul"] <- 84
begin_height["X1"][begin_height["X1"] == "May"] <- 60
begin_height["X2"][begin_height["X2"] == "Jun"] <- 72
begin_height["X2"][begin_height["X2"] == "Jul"] <- 84
begin_height["X2"][begin_height["X2"] == "May"] <- 60
begin_height$X1 <- as.numeric(as.character(begin_height$X1))
begin_height$X2 <- as.numeric(as.character(begin_height$X2))
begin_height$height <- begin_height$X1 + begin_height$X2

df_new <- cbind(s_cleaned, begin_height)
df_new %>% select(-c("X1","X2", "ht"))
df_new$height <- df_new$height %>% as.numeric() %>% replace_na(76.27)
df_new %>% select(height)
```

## Exploratory Data Analysis


```{r}
ggplot(s_cleaned, aes(x = pts)) + 
  geom_histogram()
```


Making box plots comparing the difference in the average value of certain predictors based on the player being drafted or not.
```{r}
# group by drafted or not
drafted_means <- s_cleaned %>% subset(drafted == 1) %>% sapply(mean, na.rm=TRUE)
drafted_means <- as.data.frame(drafted_means)

undrafted_means <- s_cleaned %>% subset(drafted == 0) %>% sapply(mean, na.rm=TRUE)
undrafted_means <- as.data.frame(undrafted_means)

# taking averages of drafted and undrafted players
drafted_vs_undrafted <- undrafted_means %>% mutate(drafted_means)

# transforming data frame for plotting
drafted_vs_undrafted$names <- rownames(drafted_vs_undrafted)
rownames(drafted_vs_undrafted) <- NULL
drafted_vs_undrafted <- gather(drafted_vs_undrafted, event, total, drafted_means:undrafted_means)

d_vs_u <- drafted_vs_undrafted %>% filter(names == "pts" | names == "dunksmade" |
                                            names == "min_played" | names == "ast" | 
                                            names == "TPM" | names == "midmade"| 
                                            names == "blk" | names == "total_reb" | 
                                            names == "twoPM" | names == "rimmade" |
                                            names == "stl" | names == "ht")

# plotting of means
avg_bars <- d_vs_u %>% ggplot(aes(names, total, fill=event)) + 
  geom_bar(stat = "identity", position = 'dodge') + facet_wrap(~ names, scales = "free")
avg_bars
```


Making a barplot of the percentage of players from each conference that are drafted.

We notice that there are a handful of conferences that have a drastically higher percentage of players drafted. Additionally, there are a few conferences where no players have been drafted.
```{r, fig.width=10}
draft_by_conf <-  s_cleaned %>% group_by(conf) %>% summarise(prop_drafted = mean(drafted))

ggplot(draft_by_conf) + geom_bar(aes(x = conf, y = prop_drafted), stat = "identity") +
  theme(axis.text.x = element_text(angle = 90))
```

This plot exposes a few problems with the naming of the conferences in the Data. The independent conference is listed as 2 different conferences (ind and Ind) as a result of a discrepency in the capitalization and the PAC-12 was formerly known as the PAC-10 which is why there is a P10 as well as a P12. We fix these issues below and reprint the bar graph. 
```{r}
s_cleaned["conf"][s_cleaned["conf"] == "ind"] <- "Ind"
s_cleaned["conf"][s_cleaned["conf"] == "P10"] <- "P12"
```

```{r}
draft_by_conf <-  s_cleaned %>% group_by(conf) %>% summarise(prop_drafted = mean(drafted))

ggplot(draft_by_conf) + geom_bar(aes(x = conf, y = prop_drafted), stat = "identity") +
  theme(axis.text.x = element_text(angle = 90))
```

To better understand the relationship between our predictors, we make scatteplots with 2 predictors on the axis and color indicating wheter that player was drafted.

This first scatterplot has minutes played on the x axis and Points on the y axis. As expected these two variables are highly positively correlated with players who would go on to be drafted having higer values of both ending up in the top right of this plot.
```{r}
s_cleaned %>% ggplot(aes(x=min_played, y = pts, col = drafted)) + geom_point()

```

We made another scatterplot to examine the relationship between Dunks made and three pointers made. 

```{r}
s_cleaned %>% ggplot(aes(x=dunksmade, y = TPM, col = drafted)) + geom_point()


```

# Test/training split

We will split the data into a test set and a training set. Since we have a reletively large number of observations, we should have a large enough test se if we used only 15 percent of the data for the test set, leaving many observations in the training set to fit our models.

We sample for our test set using a stratified random sample. This is a good idea in this case because 

```{r}
set.seed(123)

test_set <- s_cleaned %>% group_by(year) %>% sample_frac(size = 0.15)

test_set <- test_set[order(test_set$pid),]


training_set <- s_cleaned[!s_cleaned$pid %in% test_set$pid,]


table(test_set$drafted)
table(training_set$drafted)
```

